Beyond body length: Uncovering the drivers of vertebra number variation in Australian blindsnakes

### 2024-11-06 manuscript.Rmd
prep submit
trim abstract

The axial skeleton is important in understanding the evolution of body plans in vertebrates. Changes in the vertebral skeleton can influence body form and modes of locomotion. Elongate animals commonly show a positive correlation between body size and the number of vertebrae, a condition known as pleomerism. However, some groups of vertebrates lack pleomerism, suggesting it is possible to dissociate the production of segmental and somatic growth, but the factors that may influence this pattern are not well understood. Snakes offer a great model to study axial evolution because the number of vertebrae vary greatly between species and they exhibit a wide range of body size and shape. Here, we investigate the prevalence of pleomerism in a continental radiation of fossorial blindsnakes and test whether body shape and ecology can explain the variation in the number of vertebrae. We obtained X-ray scans and morphometric data for 349 museum specimens from 48 lineages of Australo-Papuan blindsnakes and close relatives, and evaluated their correlation with each other and ecological variables. We found an absence of pleomerism among Australian blindsnakes and a correlation between vertebra ratio and body shape, where narrow-and-long species usually possess more vertebrae for their maximum length than wide-and-robust species. We also found that species that have a greater number of vertebrae for their size are generally found in warmer environments and environments with harder-to-penetrate soil substrate. Our results extend our understanding of the axial evolution of snakes and the macroevolutionary dynamics of Australian blindsnakes.

>>>>

The axial skeleton is important in understanding the evolution of body plans in vertebrates. Changes in the vertebral skeleton can influence body form and modes of locomotion. Elongate animals commonly show a positive correlation between body size and the number of vertebrae, a condition known as pleomerism. However, some groups of vertebrates lack pleomerism, suggesting it is possible to dissociate the number of vertebrae and somatic growth, but the factors that influence this pattern are not well understood. Snakes offer a great model to study axial evolution because they exhibit a wide range of intraspecific variation in vertebra number and interspecific variation in body size, shape and ecology. Here, we investigate the prevalence of pleomerism in a continental radiation of fossorial blindsnakes and test whether body shape and ecology correlate with vertebra number. To do so, we assembled a comprehensive eco-morphological dataset including X-ray scans for 48 lineages of Australo-Papuan typhlopid snakes. We found an absence of pleomerism among Australian blindsnakes and a correlation between vertebra ratio and body shape, where narrow-and-long species usually possess more vertebrae for their maximum length than wide-and-robust species. We also found that species that have a greater number of vertebrae for their size are found in warmer environments and in harder-to-penetrate soil substrate. These results extend our understanding of the axial evolution of snakes and the macroevolutionary dynamics of Australian blindsnakes.
